---
title: "project: error types & quitting"
subtitle: "data preparation & descriptives"
output:
  html_notebook:
    theme: yeti
    toc: true
    toc_depth: 4
  html_document:
    theme: yeti
    df_print: paged
css: styles.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)
```


```{r data, message=FALSE, include=FALSE}
source(here::here("config.R"))
source(here("analysis-annie/scripts/00-setup_environment.R"))
source(here("analysis-annie/scripts/01-load_data.R"))
```

# data preparation

**The current data ranges from 1st Sept 2021 to 31st May 2024.**

I join the log data with item information from the dataset `items`. This dataset also comes with a table called `duplicate_json`. This resulted from some data checks in the item data, where some items are duplicated, but treated with different ids. This check was done in the scripts `check_items.R` and `fix_json.R`.

Then I do some initial data cleaning:

-   select only training set (training ids in folder `rdata` \> `ext`).\
-   create variables `first`, `second`, `question_clean`, `response` (factor), `comma`, `answer_id`, `single_digit`.\
-   substitute the comma for a decimal in `answer` variable, so that it can be handled numerically.

I also make the table `id_strings`, which gives, per `question_clean`, the unique ids connected to it (since some have more than 1).

## data cleaning

I remove:

-   items with fewer than `min_item_count` observations (default in `config.R`: 200)\
-   learners with fewer than `min_play_count` observations (default in `config.R`: 200)\
-   learners in grades 1 and 2

```{r data-clean, echo=FALSE}
source(here("analysis-annie/scripts/02-clean_data.R"))
```

```{r data-plot-n, fig.width=10, fig.height=5, include = TRUE, echo = FALSE}
plot_n_ind <- logs %>% 
  group_by(user_id) %>% 
  count() %>% 
  ggplot(aes(n)) + 
  geom_histogram(stat = "bin", bins = 100, color = comp_color, fill = main_color) +
  labs(title = "n. observations per individual",
       subtitle = paste(as.Date(range(logs$created)[1]),
                        "to",
                        as.Date(range(logs$created)[2]))) +
  plot_theme

# n responses per item
plot_n_item <- logs %>% 
  group_by(item_id) %>% 
  count() %>% 
  ggplot(aes(n)) + 
    geom_histogram(stat = "bin", bins = 40, color = comp_color, fill = main_color) +
    labs(title = "histogram of item count",
         subtitle = paste(as.Date(range(logs$created)[1]),
                          "to",
                          as.Date(range(logs$created)[2]))) +
    plot_theme +
    geom_histogram(data = logs %>% group_by(question_clean) %>% count(),
                   aes(n), fill = "tomato", color = "tomato3", alpha = 0.4, stat = "bin", bins = 40) +
    labs(caption = "gray bars = histogram of item_id \nred bars = histogram of question_clean")

# n per item, single-digit multiplication
plot_n_single_digit <- logs %>% 
  filter(single_digit) %>% 
  group_by(question_clean, first, second) %>% 
  count() %>% 
  ungroup() %>% 
  ggplot(aes(first, second, fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  #geom_text(aes(label = n), color = "white", size = 3) +
  scale_x_continuous(breaks = 0:9, name = "First operand") +
  scale_y_reverse(breaks = 0:9, name = "Second operand") +
  scale_fill_viridis_c(option = "rocket", end = 0.05, begin = 0.9, name = "n") +
  coord_equal() +
  labs(title = "n. observations",
  subtitle = "single-digit multiplication items") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0, face = "bold")
  )


plot_grid(plot_n_single_digit, plot_n_item)
```

## response time variables

I create the variables:

-   `rt_fastslow` in the log data. On a log level, was the response slower or faster than the median response time?\
-   On an item level (grouped by `question_clean`):
-   `rt_totalmean`: average response time per item.\
-   `rt_totalsd`: standard deviation of response times per item.\
-   `rt_mean`: average response time, grouped by item and response type (correct, incorrect, skip).
-   `rt_sd`: standard deviation of response times, grouped by item and response type (correct, incorrect, skip).

```{r vars-rt}
rt %>% 
  filter(single_digit)

logs %>% 
  filter(response != "late") %>% 
  group_by(response) %>% 
  summarise(mean = mean(response_in_milliseconds),
            median = median(response_in_milliseconds),
            sd = sd(response_in_milliseconds))
```

### `totals`

-   grouped by item
-   n of obs per response type
-   error rate = `n_error / (n_cor + n_error)`

```{r vars-totals}
totals %>% 
  filter(single_digit)
```
### `resp`

-   all responses, grouped by answer

```{r vars-resp}
resp %>% 
  filter(question_clean == "3 x 3") %>% 
  arrange(-n)
```
