---
title: "project: error types & quitting"
subtitle: "data preparation & descriptives"
output:
  html_document:
    df_print: paged
  html_notebook:
    theme: yeti
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)
```


```{r data, message=FALSE, include=FALSE}
source(here::here("config.R"))
source(here("analysis-annie/scripts/00-setup_environment.R"))
source(here("analysis-annie/scripts/01-load_data.R"))
```

# data preparation

**The current data ranges from 1st Sept 2021 to 31st May 2024.**

I join the log data with item information from the dataset `items`. This dataset also comes with a table called `duplicate_json`. This resulted from some data checks in the item data, where some items are duplicated, but treated with different ids. This check was done in the scripts `check_items.R` and `fix_json.R`.

Then I do some initial data cleaning:

-   select only training set (training ids in folder `rdata` \> `ext`).\
-   create variables `first`, `second`, `question_clean`, `response` (factor), `comma`, `answer_id`, `single_digit`.\
-   substitute the comma for a decimal in `answer` variable, so that it can be handled numerically.

I also make the table `id_strings`, which gives, per `question_clean`, the unique ids connected to it (since some have more than 1).

## data cleaning

I remove:

-   items with fewer than `min_item_count` observations (default in `config.R`: 200)\
-   learners with fewer than `min_play_count` observations (default in `config.R`: 200)\
-   learners in grades 1 and 2

```{r data-clean, message=FALSE, include=FALSE}
source(here("analysis-annie/scripts/02-clean_data.R"))
```

```{r data-plot-n, fig.width=10, fig.height=5, include = TRUE, echo = FALSE}
plot_n_ind <- logs %>% 
  group_by(user_id) %>% 
  count() %>% 
  ggplot(aes(n)) + 
  geom_histogram(stat = "bin", bins = 100, color = comp_color, fill = main_color) +
  labs(title = "n. observations per individual",
       subtitle = paste(as.Date(range(logs$created)[1]),
                        "to",
                        as.Date(range(logs$created)[2]))) +
  plot_theme

# n responses per item
plot_n_item <- logs %>% 
  group_by(item_id) %>% 
  count() %>% 
  ggplot(aes(n)) + 
    geom_histogram(stat = "bin", bins = 40, color = comp_color, fill = main_color) +
    labs(title = "histogram of item count",
         subtitle = paste(as.Date(range(logs$created)[1]),
                          "to",
                          as.Date(range(logs$created)[2]))) +
    plot_theme +
    geom_histogram(data = logs %>% group_by(question_clean) %>% count(),
                   aes(n), fill = "tomato", color = "tomato3", alpha = 0.4, stat = "bin", bins = 40) +
    labs(caption = "gray bars = histogram of item_id \nred bars = histogram of question_clean")

# n per item, single-digit multiplication
plot_n_single_digit <- logs %>% 
  filter(single_digit) %>% 
  group_by(question_clean, first, second) %>% 
  count() %>% 
  ungroup() %>% 
  ggplot(aes(first, second, fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  #geom_text(aes(label = n), color = "white", size = 3) +
  scale_x_continuous(breaks = 0:9, name = "First operand") +
  scale_y_reverse(breaks = 0:9, name = "Second operand") +
  scale_fill_viridis_c(option = "rocket", end = 0.05, begin = 0.9, name = "n") +
  coord_equal() +
  labs(title = "n. observations",
  subtitle = "single-digit multiplication items") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0, face = "bold")
  )


plot_grid(plot_n_single_digit, plot_n_item)
```

## response time variables

I create the variables:

-   `rt_fastslow` in the log data. On a log level, was the response slower or faster than the median response time?\
-   On an item level (grouped by `question_clean`):
-   `rt_totalmean`: average response time per item.\
-   `rt_totalsd`: standard deviation of response times per item.\
-   `rt_mean`: average response time, grouped by item and response type (correct, incorrect, skip).
-   `rt_sd`: standard deviation of response times, grouped by item and response type (correct, incorrect, skip).

```{r vars-rt}
rt %>% 
  filter(single_digit)

logs %>% 
  filter(response != "late") %>% 
  group_by(response) %>% 
  summarise(mean = mean(response_in_milliseconds),
            median = median(response_in_milliseconds),
            sd = sd(response_in_milliseconds))
```

### `totals`

-   grouped by item
-   n of obs per response type
-   error rate = `n_error / (n_cor + n_error)`

```{r vars-totals}
totals %>% 
  filter(single_digit)
```
### `resp`

-   all responses, grouped by answer

```{r vars-resp}
resp %>% 
  filter(question_clean == "3 x 3") %>% 
  arrange(-n)
```

# most common errors

### `top_errors`

-   include only incorrect responses
-   group by answer\
-   aggregate n per unique answer for top 10 rows per item

For example, for the item `9 x 9`:

```{r top-errors, include = TRUE, echo = FALSE}
top_errors <- resp %>% 
  filter(response != "cor") %>% 
  group_by(question_clean) %>% 
  arrange(question_clean, desc(n)) %>% 
  slice_head(n = 30) %>%
  ungroup() %>%
  arrange(first, second, desc(n)) %>% 
  mutate(prop_error = ifelse(response != "error", NA, n / n_error)) 
  
top_errors_fastslow <- resp_fastslow %>% 
  filter(response != "cor") %>% 
  group_by(question_clean, rt_fastslow) %>% 
  arrange(question_clean, desc(n)) %>% 
  slice_head(n = 60) %>%
  ungroup() %>%
  arrange(question_clean, first, second, desc(n)) %>% 
  mutate(prop_error = ifelse(response != "error", NA, n / n_error)) 
```

```{r errors-top-ten-1, fig.width = 8, fig.height = 10, include = TRUE, echo = FALSE}
top_ten_errors <- top_errors %>% 
  filter(filter_error & single_digit) %>%
  group_by(question_clean) %>% 
  slice_head(n = 1) %>% 
  arrange(desc(n)) %>%
  ungroup() %>% 
  slice_head(n = 10) %>%
  mutate(line = paste0(answer_id, " (n = ", n, ")")) %>%
  pull(line)
```

**The top 10 most common errors among single-digit multiplication are:**

```{r errors-top-ten-2, include = TRUE, echo = FALSE}
cat(top_ten_errors, sep = "\n")
```

```{r common-errors-plot, include = TRUE, echo = FALSE, fig.width = 12, fig.height=10}
select_items <- top_errors %>% 
  filter(single_digit & filter_error & !first %in% 0 & !second %in% 0) %>% 
  group_by(question_clean) %>% 
  arrange(question_clean, desc(prop_error)) %>% 
  slice_head(n = 1) %>% 
  arrange(desc(prop_error)) %>% 
  filter(n > 50) %>% # same error has been recorded more than 50x
  pull(question_clean) %>% 
  .[1:15] # preferably a factor of 5

top_errors %>% 
  # filter single-digit, "true" error, and no zeroes in item. 
  filter(single_digit & filter_error & !first %in% 0 & !second %in% 0) %>% 
  group_by(question_clean) %>% 
  arrange(question_clean, desc(prop_error)) %>% 
  # remove gibberish long answers 
  mutate(answer_str = as.character(answer)) %>% 
  filter(str_length(answer_str) < 6) %>%
  # sub selection of items
  filter(question_clean %in% select_items) %>% 
  arrange(question_clean, -prop_error) %>%
  group_by(question_clean) %>% 
  slice_head(n = 15) %>% 
  ungroup() %>% 
  mutate(answer = reorder_within(answer, prop_error, question_clean)) %>% 
  ggplot(aes(answer, prop_error)) +
  geom_bar(stat = "identity", aes(fill = response, color = response), alpha = 0.4) +
  scale_discrete_manual(aesthetics = c("fill", "color"), values = response_colors, name = "type") +
  labs(title = "Number of responses per error", y = "n", x = "answer") +
  coord_flip() +
  facet_wrap(~question_clean, scales = "free_y", ncol = 5) + 
  scale_x_reordered() +  # clean up facet x labels
  plot_theme
```

## entropy of error distributions

To quantify how concentrated or diffuse the error distributions are for each item, I compute the **Shannon entropy** of the item-specific error distribution.

For each item, the empirical distribution of incorrect answers is considered (plotted above).

The entropy of an item is defined as

$H = -\sum_{k=1}^{K} p_k \log p_k$

-   $p_k$ denotes the proportion of errors resulting in answer $k$.

-   $K$ denotes the number of distinct incorrect answers.

### normalization

Because the maximum possible entropy depends on $K$, raw entropy values are not directly comparable across items. The maximum entropy for an item with $K$ distinct errors is $\log(K)$, which occurs when all incorrect answers are equally likely. I therefore normalize entropy by this maximum:

$H_{norm} = \frac{H}{\log(K)}$

This normalized entropy takes on values between 0 and 1, where 0 indicates that all errors fall on a single answer and 1 indicates a uniform distribution over the observed error answers.

### dominance

For interpretation, I compute the inverse entropy measure, referred to as **dominance**:

$
\mathrm{dominance} = 1 - H_{norm}
$

High dominance values correspond to sharply peaked error distributions, where one incorrect error dominates. Low dominance values correspond to flatter distributions with many competing error answers.

### note:

-   Only incorrect answers are included in the entropy calculation. *Skipped and late responses are not included*.

-   Proportions (`prop_error = n / n_error`) are derived from the observed error counts for each item. Thus, probabilities used in the entropy calculation depend on relative frequencies. Indirectly, this implies that, for items with a higher total play count, these proportions are more stable, compared to items with a relatively smaller play count. Items with n \< 200 are excluded from analysis.


```{r entropy-calculate, include = TRUE}
errors <- resp %>% 
  filter(filter_error) %>% 
  arrange(first, second, desc(n)) %>% 
  mutate(prop_error = ifelse(response != "error", NA, n / n_error)) 

entropy <- errors %>%
  filter(filter_error) %>%
  group_by(question_clean) %>%
  summarise(
    single_digit = unique(single_digit),
    first = unique(first), 
    second = unique(second), 
    H = -sum(prop_error * log(prop_error), na.rm = TRUE),
    K = n(),
    H_norm = H / log(K),
    dominance = 1 - H_norm,
    .groups = "drop"
  )

# highest dominance items 
entropy %>% 
  filter(single_digit) %>% 
  arrange(-dominance) %>% 
  slice_head(n = 12) %>% 
  dplyr::select(question_clean, H, K, H_norm, dominance)
```

```{r entropy-join-item-info, }
entropy <- entropy %>% 
  left_join(items %>% dplyr::select(question_clean, rating), by = "question_clean") %>%
  left_join(totals %>% dplyr::select(question_clean, n_qm, n_late, n_error), by = "question_clean") %>% 
  # remove duplicates
  distinct(question_clean, .keep_all = TRUE)
```

```{r entropy-plot, include = TRUE, echo = FALSE, fig.width = 10, fig.height=6}
entropy_hist <- entropy %>% 
  ggplot() +
  geom_histogram(aes(x = dominance, color = single_digit, fill = single_digit), 
                 bins = 50, 
                 alpha = 0.6) +
  geom_histogram(data = entropy, 
                 aes(x = dominance, color = single_digit, fill = single_digit),
                 bins = 50, 
                 alpha = 0.6) +
  labs(title = "Histogram of Item Dominance",
       x = "Dominance (Inverse Entropy)", 
       subtitle = "Higher values = more peaked distribution",
       color = "single-digit", 
       fill = "single-digit") +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.4) +
  scale_fill_viridis_d(option = "magma", begin = 0.8, end = 0.4) +
  plot_theme

# entropy per item, single-digit multiplication
plot_entropy_singledigit <- entropy %>% 
  filter(first != 0 & second != 0 & single_digit) %>% 
  ggplot(aes(first, second, fill = dominance)) +
  geom_tile(color = "white", linewidth = 0.3) +
  #geom_text(aes(label = n), color = "white", size = 3) +
  scale_x_continuous(breaks = 0:9, name = "First operand") +
  scale_y_reverse(breaks = 0:9, name = "Second operand") +
  scale_fill_viridis_c(option = "rocket", end = 0.05, begin = 0.9, name = "n") +
  coord_equal() +
  labs(title = "dominance",
  subtitle = "single-digit multiplication items",
  caption = "0 table is removed; these have very high dominance values (>0.7)") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0, face = "bold")
  )

plot_grid(entropy_hist, plot_entropy_singledigit, rel_widths = c(0.8, 1))
```


Looking at response distributions for high- and low-entropy items:

### plot individual error distributions

```{r entropy-ind-items-plot, include = TRUE, echo = FALSE, fig.width=12, fig.height=6}
source(here("analysis-annie/functions/entropy-functions.R"))
plot_high_dom <- plot_entropy_items(
  only_single_digit = TRUE,
  n_items = 6,
  entropy, top_errors, response_colors,
  decreasing = TRUE,
  subtitle_text = "Highest dominance items"
)

plot_low_dom <- plot_entropy_items(
  only_single_digit = TRUE,
  n_items = 6,
  entropy, top_errors, response_colors,
  decreasing = FALSE,
  subtitle_text = "Lowest dominance items"
)

plot_grid(plot_high_dom, plot_low_dom)
```

### entropy and item rating

Here I am looking at the relationship between entropy and item rating.

**Some observations:**

-   There is a clear positive association between rating and entropy: items with more dominance (more peaked distributions) tend to have lower rating estimates. In short, more difficult items elicit more structured error distrubtions.
    -   **interpretation:** errors on easier items are often driven by a small number of salient, rule-based misconceptions which leads to strongly concentrated error distributions. because these items are otherwise easy, errors are rare but when they do occur, they tend to collapse onto a single dominant incorrect response. in contrast, more difficult items admit a larger space of plausible but incorrect strategies. this does not imply that difficult items are less informative, rather that no single erroneous procedure dominates across learners.
-   There a positive association between the absolute number of incorrect responses and dominance: items with higher dominance tend to accumulate a larger number of errors overall.
    -   interpretation: this likely reflects the adaptivity of the system. easier items are played more, and these also show higher dominance.

```{r, include=TRUE, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE}
plot_entropy_rat <- entropy %>% 
  filter(dominance < 0.5) %>% 
  ggplot(aes(dominance, rating)) +
  geom_point(color = comp_color, aes(alpha = single_digit)) +
  #geom_smooth(method = "lm", se = F, color = comp_color) +
  plot_theme

plot_entropy_qm <- entropy %>% 
  filter(dominance < 0.5) %>% 
  ggplot(aes(dominance, n_qm)) +
  geom_point(aes(alpha = single_digit), color = response_colors["qm"]) +
  #geom_smooth(method = "lm", se = F, color = response_colors["qm"]) +
  plot_theme + 
  theme(legend.position = "none")

plot_entropy_oot <- entropy %>% 
  filter(dominance < 0.5) %>% 
  ggplot(aes(dominance, n_late)) +
  geom_point(aes(alpha = single_digit), color = response_colors["late"]) +
  #geom_smooth(method = "lm", se = F, color = response_colors["late"]) +
  plot_theme + 
  theme(legend.position = "none")

plot_entropy_error <- entropy %>% 
  filter(dominance < 0.5) %>% 
  ggplot(aes(dominance, n_error)) +
  geom_point(aes(alpha = single_digit), color = response_colors["error"]) +
  #geom_smooth(method = "lm", se = F, color = response_colors["error"]) +
  plot_theme + 
  theme(legend.position = "none")

plot_grid(plot_entropy_rat, plot_grid(plot_entropy_qm, plot_entropy_oot, plot_entropy_error, ncol = 1))

entropy %>% arrange(-rating) %>% dplyr::select(question_clean, rating)
```

### entropy and response time

```{r entropy-rt, }
# create quantiles of dominance scores 
n_quant = 10
entropy <- entropy %>% 
  mutate(dom_quantile = factor(ntile(dominance, n_quant)))

groupsize <- entropy %>% 
  group_by(dom_quantile) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(mean = mean(n)) %>% 
  pull(mean)
```

I separate dominance into quantiles, and then examine response time distributions of items across these quantiles. I also plot average and standard deviations of response times across items as a function of dominance. 

```{r entropy-rt-plot, include=TRUE, echo=FALSE, fig.width = 12, fig.height=8, warning=FALSE}
cat("n quantiles = ", n_quant, "(average groupsize = ", groupsize, ")")

plot_dom_density <- logs %>% 
  left_join(entropy %>% dplyr::select(question_clean, dominance, dom_quantile), by = "question_clean") %>% 
  filter(response == "error" & 
           question_clean %in% unique(entropy$question_clean)) %>% 
  # filter(dominance < 0.5) %>%
  # # peaked distribution of rts is mostly driven by outlier group 
  ggplot() + 
  geom_density(aes(x = response_in_milliseconds, 
                   fill = dom_quantile, 
                   color = dom_quantile),
               stat = "density", alpha = 0.2) + 
  scale_fill_viridis_d(option = "magma") +
  scale_color_viridis_d(option = "magma") +
  plot_theme +
  labs(title = "rt distributions per dominance quantile",
       subtitle = paste("average group size =", round(groupsize, 2)),
       fill = "quantile \n(lowest to highest \ndominance)",
       color = "quantile \n(lowest to highest \ndominance)")
  
plot_dom_meanrt <- logs %>% 
  left_join(entropy %>% dplyr::select(question_clean, dominance, dom_quantile), by = "question_clean") %>% 
  filter(response == "error" & 
           question_clean %in% unique(entropy$question_clean)) %>% 
  group_by(question_clean) %>% 
  summarise(single_digit = unique(single_digit), 
            mean_rt = mean(response_in_milliseconds), 
            dom = unique(dominance)) %>% 
  ggplot(aes(x = dom, y = mean_rt)) +
  geom_point(aes(color = single_digit), alpha = 0.4) + 
  #geom_smooth(method = "lm", se = FALSE, color = "gray20") +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.4) +
  plot_theme + 
  labs(x = "dominance", 
       y = "mean rt (ms)", 
       color = "single-digit")

plot_dom_sdrt <- logs %>% 
  left_join(entropy %>% dplyr::select(question_clean, dominance, dom_quantile), by = "question_clean") %>% 
  filter(response == "error" & 
           question_clean %in% unique(entropy$question_clean)) %>% 
  group_by(question_clean) %>% 
  summarise(single_digit = unique(single_digit), 
            sd_rt = sd(response_in_milliseconds), 
            dom = unique(dominance)) %>% 
  ggplot(aes(x = dom, y = sd_rt)) +
  geom_point(aes(color = single_digit), alpha = 0.4) + 
  #geom_smooth(method = "lm", se = FALSE, color = "gray20") +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.4) +
  plot_theme +
    labs(x = "dominance", 
       y = "standard deviation of rt (ms)", 
       color = "single-digit")

plot_grid(plot_dom_density, plot_grid(plot_dom_meanrt + theme(legend.position = "none"), plot_dom_sdrt, ncol = 1, rel_heights = c(1, 1.1)), rel_widths = c(1, 0.8))
```

ideas for entropy:

-   compute entropy separately by age group for the same item

    -   possible relationships to look for: decreasing entropy with age (consistent with consolidation of error procedures) ; persistence of high dominance across ages (robust misconception?) ; disappearance of dominance with age (developmental phase error?)

    -   interactions with rt?

-   compute person-specific (or day-specific) entropy across items?

